---
title: "Extracting and analyzing data from Google Earth Engine using rgee"
author: "Jelany Duali"
date: "2024-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A look at temperature at land surface temperature in Guelph, Ontario

Here, I briefly illustrate how you can extract data from  Google Earth Engine using rgee in R. Since there is a lot of talk about rising temperatures, I have decided to work with temperature data measured in the city of Guelph. 

I chose Guelph because it is a small area, so processing times for the data won't be too long.

I'll start by loading some libraries (including rgee) and setting my working directory  

```{r, echo=T, results = "hide", message=F}

# load libraries
lapply(X = c("terra", "tidyverse", "sf", "anytime", "rgee", "viridis"), FUN = library, 
       character.only = TRUE)

# set working directory
setwd("C:/Users/Jelan/OneDrive/Desktop/University/University of Guelph/Other/Remote_sensing_seminar/rgee_examples/rgee_seminar_code")

```

### Initializing rgee

Before we start using rgee, we need to authorize it to manage Earth Engine, Google Drive, and Google Cloud Storage through my Google account. 

This process is usually a bit simpler than what is shown below. I think there is currently an issue related to the ee$Initialize() function and I have to work around it. 

```{r, echo=T}
# Set python environment
earthengine_python <- Sys.getenv("EARTHENGINE_PYTHON", unset = NA)
Sys.setenv(RETICULATE_PYTHON = earthengine_python)

# Initialize rgee and check credentials, this requires a bit more work because ee$Initialize() malfunctions
ee$Authenticate(auth_mode='notebook')
ee$Initialize(project='jelany-project-data')

# Check that everything is working properly
ee_check()

```

Next we want to load our spatial datasets. For temperature we will use the MOD11A1.061 dataset, which we will extract as an image collection.

This dataset is available from the Google Earth Calendar: https://developers.google.com/earth-engine/datasets/catalog/MODIS_061_MOD11A1 

We will also need a polygon outlining the city of Guelph, we will use a simple polygon dataset available here: 
https://www.arcgis.com/apps/mapviewer/index.html?layers=6edf63b03c95423bae1ab9c590866838 

This is actually a set of polygons for each property in Guelph, but we will merge those into a single polygon. As you can tell from looking at the output, there are a few artefacts inside our polygon, but that's fine. 

We will use the Guelph city polygon as a "mask" to filter through our dataset of land surface tempetature (LST). This is because we only need data for the city of Guelph. 

```{r, echo=T, message=F}

# We extract the data from the GEE catalog as an image collection 
temp.gee <- ee$ImageCollection("MODIS/061/MOD11A1")

# I downloaded the Guelph city polygon dataset, so I load it and merge the polygons. 
# At the end, I convert the polygon from the sf format to an ee.feature  object 
sf_use_s2(FALSE)
mask <- st_read("Property/Property.shp") 
g.crs <-crs(mask) 
mask <- mask %>%
  st_as_sf() %>%
  st_union() %>%
  sf_as_ee()

#Let's have a look at our guelph city polygon
Map$setCenter(-80.24, 43.54, 10)
Map$addLayer(mask)

# Let's apply the mask to our spatial dataset. We'll discard all the data in there except the one in Guelph 
# Note that we are using functions from the Earth Engine server 
# This is because we are working with an Earth Engine server object 
temp.gee <- temp.gee$map(function(img){
  img <- img$clip(mask)
})

```

### What was the mean temperature in Guelph in 2023?

Now that we have all our data, we can start working with it. For example, let's have a look at the mean temperature in Guelph in 2023 

```{r, echo=T, message=F}

# We filter our MODIS feature collection and extract the mean temperature in 2023
temp23 <- temp.gee$filter(ee$Filter$date('2023-01-01','2023-12-31'))$select('LST_Day_1km')$mean()

# Let's visualize this 
# we create a palette first 
temp23.palette <- list(
  min = 13000,
  max = 16500,
  palette = turbo(100)
)

# Then we add the image onto a map 
Map$setCenter(-80.24, 43.54, 10)
Map$addLayer(temp23, temp23.palette)

```

It is a bit hard to make sense of our the image produced here, especially since there is no scale bar. 

We can convert to image to a file and export it to Google drive (In this case I save it in my own account).

Note that when we plot the data now, we are no longer using functions from Earth Engine. The scale of the MODIS/061/MOD11A1 dataset is 1x1 km, so this is the size of our pixels. 

```{r, echo=T, message=F}

# We can convert the image we created to a file, we'll export it to Google Drive
task <- ee_image_to_drive(
  image = temp23,
  scale = 1000, # resolution of 1 pixel per 10 kilometers
  folder= "GEE_imports",
  fileFormat = "GEO_TIFF",
  region = mask)

# task$start()
# ee_monitoring()

# Now we can read this file with the terra package
guelph.temp23 <- terra::rast("rgee_guelph_temp.tif")
plot(guelph.temp23)

# the temperature here is in Kelvin, and scaled by 0.02

#Let's rescale the temperature to see it in Celsius 
guelph.temp23 <- guelph.temp23*0.02- 273.15
plot(guelph.temp23)

```

It looks like downtown Guelph is a small heat island. Let's move on to a more advanced analysis 

### Guelph's temperature over time 

How has Guelph's temperature changed overtime since 2001? Here, I'll extract Guelph's monthly temperature for every month from 2001 to 2023. 

Recall that we are working with an image collection. We'll iterate through every month and year, and for each month, we will filter our collection to get all image for the month. Then, we will extract the mean temperature for the month by reducing all images in that sub collection. We will store our monthly mean data, as well as year and month in a feature collection.  

This process is a but similar to a nested loop in R. We loop through each year, and through each month in a year. 

```{r, echo=T, message=F}

# The years and month that we will be iterating through 
years <- ee$List$sequence(2001 , 2023)
months <- ee$List$sequence(01 , 12)

# let's extract the data 
monthly.temp <- ee$FeatureCollection(years$map(ee_utils_pyfunc(function(year){
  return(months$map(ee_utils_pyfunc(function(month){
    
    monthly.img <- temp.gee$filter(ee$Filter$calendarRange(year, year, "year"))$
             filter(ee$Filter$calendarRange(month, month, "month"))$
             select('LST_Day_1km')$mean()
    
    monthly.temp <- monthly.img$clip(mask)$reduceRegion(reducer = ee$Reducer$mean())
    
    return(ee$Feature(NULL, list("month_temp" = monthly.temp$get("LST_Day_1km"), 
                                 'year' = year, 
                                 "month" = month)))
    
  })))
}))$flatten())

```

Again, I'll export the output to Google Drive, download it, and load it in R

```{r, echo=T, message=F}

#export the output to Google Drive
# By default, only properties that are non-NUll in the first image are exported. They might not include the output.
task <- ee_table_to_drive(
  collection = monthly.temp,
  description = "monthly_temp_mean_canada",
  folder = "GEE_imports",
  fileFormat = "CSV")

#task$start()
#ee_monitoring()

# Let's load our data and have a look a it 
monthly.temp <- read.csv("monthly_temp_mean_Guelph.csv")
head(monthly.temp)
```

Now, we can plot our temperature data using ggplot. I'll just do a bit of work first so we have a date for each data point, and to convert our measurements from Kelvin to Celcius

```{r, echo=T, message=F}

# Add a column with dates (every monthly mean is set to the 1st of the month)
monthly.temp$Date <- anytime(paste0(monthly.temp$year, "-", monthly.temp$month))

# Convert from Kelvin to Celcius
monthly.temp$month_temp <- monthly.temp$month_temp*0.02- 273.15

# We create our plot
ggplot(data = monthly.temp, aes(y = month_temp, x = Date)) + 
  geom_line()+
  geom_smooth(method='lm', formula= y~x)+
  labs(x = "Date", y = "Temperature")

```

We can also use the stl function (available in Base R) to decompose our time series into seasonal, trend and irregular components.

```{r, echo=T, message=F}

temp.ts <- ts(data = monthly.temp$month_temp, start = c(2001, 1), end = c(2023,12), deltat = 1/12) 
temp.comp <- stl(temp.ts, s.window = 'periodic')

plot(temp.comp)

```
